---
title: '1. xQTLbiolinks: query and download'
author: "RuoFan Daing"
date: "2023-05-03"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{query_download}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
lang: en-US
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo=TRUE,
  progress =FALSE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Load packages:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=TRUE}
library(xQTLbiolinks)
library(data.table)
library(stringr)
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
temp1 <- tempfile(fileext=".zip")
download.file("https://github.com/dingruofan/exampleData/raw/master/query_download_vignette.zip", temp1)
load(unz(temp1,"query_download_vignette.RData"))
close(file(temp1))
rm(temp1)
```

## xQTL Query:

### Query eQTLs:

####  Query significant eQTL associations with a variant id across all tissues:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
eQTL_sig1 <- xQTLquery_eqtlSig("rs201327123")
eQTL_sig1 <- xQTLquery_eqtlSig("chr1_14677_G_A_b38")
# Query significant eQTL associations with a variant id in a specified tissue:
eQTL_sig1 <- xQTLquery_eqtlSig("chr1_14677_G_A_b38",
                    tissueSiteDetail="Skin - Sun Exposed (Lower leg)")
```
```{r, eval=TRUE}
eQTL_sig1
```
                   
#### Query eQTL associations for multiple variants:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
varInfo <-  xQTLquery_varPos(chrom="chr1", pos=c(1102708))
eQTL_sig2 <- xQTLquery_eqtlSig(variantName=varInfo$snpId)
```
```{r, eval=TRUE}
head(eQTL_sig2)
```

#### Query eQTL associations by genes or tissues:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
eQTL_sig3 <- xQTLquery_eqtlSig(genes="ATAD3B")
eQTL_sig3 <- xQTLquery_eqtlSig(genes=c("TP53", "SLC35E2B"), tissueSiteDetail= "Brain - Cerebellum")
eQTL_sig3 <- xQTLquery_eqtlSig(genes="ENSG00000141510.16")
```
```{r, eval=TRUE}
head(eQTL_sig3)
```

#### Query eQTL associations with a variant-gene pair:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
eQTL_sig4 <- xQTLquery_eqtlSig(variantName="rs1641513", genes="TP53")
eQTL_sig4 <- xQTLquery_eqtlSig(variantName="chr1_1667948_A_G_b38",
                     genes="SLC35E2B", tissueSiteDetail="Kidney - Cortex")
```
```{r, eval=TRUE}
eQTL_sig4
```

Multi-tissue eQTL metasoft results for a given gene and variant can be also queried using `xQTLquery_eqtl`, the results include: m-value (mValue), normalized effect size (nes), p-value (pValue) and standard error (se).
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
# Query with a gene symbol:
eqtlMeta <- xQTLquery_eqtl(gene="TP53")
# Query with a variant-gene pair:
eqtlMeta <- xQTLquery_eqtl(variantName="rs1641513",gene="TP53")
```
```{r, eval=TRUE}
head(eqtlMeta)
```

### Query sQTLs:

#### Query sQTL associations with rsid:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
sQTL_sig1 <- xQTLquery_sqtlSig(variantName="rs201327123")
sQTL_sig1 <- xQTLquery_sqtlSig(variantName="chr1_14677_G_A_b38", tissueSiteDetail="Whole Blood")
```
```{r, eval=TRUE}
sQTL_sig1
```

#### Query sQTL associations with gene symbol and gencode ID:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
sQTL_sig2 <- xQTLquery_sqtlSig(genes="ENSG00000141510.16", tissueSiteDetail="Lung" )
sQTL_sig2 <- xQTLquery_sqtlSig(genes=c("ATAD3B", "MLH1"))
```
```{r, eval=TRUE}
sQTL_sig2
```

#### Query sQTL associations with the variant-genes pair:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
sQTL_sig3 <- xQTLquery_sqtlSig(variantName="rs201327123", genes=c("WASH7P","RP11-206L10.2"))
sQTL_sig3 <- xQTLquery_sqtlSig(variantName="chr17_7465085_A_G_b38", genes="TP53", 
                               tissueSiteDetail="Lung")
```
```{r, eval=TRUE}
sQTL_sig3
```

## xQTLs Download
### eQTL Download:
#### Download all eQTL associations for MLH1-rs13315355 pair in all tissues from all studies:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
eqtlAssos <- xQTLdownload_eqtlAllAsso(gene="MLH1", variantName = "rs13315355", study="")
```

#### Download all eQTL associations for gene ATP11B in Muscle - Skeletal from GTEx_V8:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
eqtlAssos <- xQTLdownload_eqtlAllAsso("ATP11B", tissueLabel="Muscle - Skeletal")
```

#### Download all eQTL associations for SNP rs11568818 in all tissues from all supported studies:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
eqtlAssos <- xQTLdownload_eqtlAllAsso(variantName="rs11568818", study="")
```

#### Download all eQTL associations of SNP rs11568818 in Muscle - Skeletal from GTEx_V8:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
eqtlAssos <- xQTLdownload_eqtlAllAsso(variantName="chr11_102530930_T_C_b38",
                                   tissueLabel="Muscle - Skeletal", study="GTEx_V8")
```

#### Download all eQTL associations for gene ATP11B in CD4+ T cell from all supported studies:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
eqtlAssos <- xQTLdownload_eqtlAllAsso(gene="MMP7",tissueLabel = "CD4+ T cell", study="")
```

```{r, eval=TRUE}
eqtlAssos
```

### sQTL Download:
#### Download all sQTL associations for gene MMP7 and TP53 in Lung:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
sqtlAssos <- xQTLdownload_sqtlAllAsso(genes=c("MMP7","TP53"), tissue="Lung")
```
```{r, eval=TRUE}
sqtlAssos
```

### 3'aQTL Download:
#### Download all 3'aQTL associations for gene MMP7 in Lung:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
aqtlAssos <- xQTLdownload_xqtlAllAsso(genes=c("MMP7"), tissue="Lung")
```
```{r, eval=TRUE}
aqtlAssos
```


## Query details of gene, variant, tissue and samples in GTEx:

#### Query gene details with gene symbols, versioned or unversioned gencode ID:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
geneInfo <- xQTLquery_gene("TP53")
geneInfo <- xQTLquery_gene(c("tp53","naDK","SDF4") )
geneInfo <- xQTLquery_gene(c("ENSG00000210195.2","ENSG00000078808"))
```
```{r, eval=TRUE}
geneInfo
```

#### Query variants using dbSNP ID or variant ID:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
varInfo <- xQTLquery_varId("rs12596338")
varInfo <- xQTLquery_varId("chr11_66561248_T_C_b38")
varInfo <- xQTLquery_varPos(chrom="chr1", pos=c(1102708,1105739))
```
```{r, eval=TRUE}
varInfo
```

#### Query tissue using tissue name:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
tissueAll <- xQTLquery_tissue() # fetch all tissues in GTEx
Brain <- xQTLquery_tissue("Brain")
```
```{r, eval=TRUE}
Brain
```


#### Query sample using tissue name or sample ID:
```{r, results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
sampleInfo <- xQTLquery_sampleByTissue("Brain - Amygdala" )
sampleInfo <- xQTLquery_sampleByTissue(tissueSiteDetail="Liver", pathologyNotesCategories=TRUE)
sampleIds <- c("GTEX-11NUK-0011-R4a-SM-DO12B", "GTEX-11ONC-0011-R4b-SM-DO93H",
               "GTEX-11DXY-0526-SM-5EGGQ", "GTEX-13OVJ-1026-SM-5IFGI")
sampleInfo <- xQTLquery_sampleBySampleId(sampleIds)
```
```{r, eval=TRUE}
sampleInfo
```




